---
title: "Week 1 Data Lab: Loading and Visualizing Data Structures"
format:
  html:
    code-fold: false
---

::: {.callout-info title="Lab Overview"}
In this practical lab, we will load different data tables and visualize their structures as discussed in the theoretical session. We will explore how Cross-Sectional, Time Series, and Panel data look in R.
:::

## Setting up the environment

We start by loading the `tidyverse` package, which provides a cohesive set of tools for data manipulation and visualization.

```{r setup, message=FALSE, warning=FALSE}
# Load necessary libraries
library(tidyverse)
```

## 1. Cross-Sectional Data

Cross-sectional data are observations of a sample of elements taken at one specific point in time. We will use the `mtcars` dataset built into R.

```{r}
# Load cross-sectional data
data(mtcars)

# Visualize structure
glimpse(mtcars)

# Show first few rows
head(mtcars)
```

## 2. Time Series Data

Time series data consist of observations of one or more variables over time. Here we use the `economics` dataset from ggplot2.

```{r}
# Load time series data
data(economics)

# Visualize structure
glimpse(economics)

# Plotting time series structure quickly
ggplot(economics, aes(x = date, y = psavert)) +
  geom_line() +
  labs(title = "Personal Savings Rate Over Time", x = "Date", y = "Savings Rate")
```

## 3. Panel or Longitudinal Data

Panel data follows the same elements over multiple time periods. We can simulate a simple structure to understand how it organizes individuals across time.

```{r}
# Creating a simple mock panel dataset
set.seed(123)
panel_data <- tibble(
  individual_id = rep(1:5, each = 3),
  year = rep(2021:2023, times = 5),
  income = round(runif(15, 30000, 80000), 0)
)

# Visualize structure
glimpse(panel_data)

# Show first few rows
head(panel_data)
```

# Session Information

Providing the session information ensures reproducibility by recording the exact versions of R and packages used in this analysis.

```{r session-info}
# Display session info
sessionInfo()
```

::: {.callout-tip title="Cross-References"}
* Return to the [Week 1 Theory](../../Sessions/W1_Intro_and_data/index.qmd)
* Return to the [Course Root](../../index.qmd)
:::
