---
title: "Week 1: Introduction and Data Structures"
format:
  html:
    code-fold: false
---

> [!info] Overview
>
> This session introduces the foundational concepts of quantitative methods and the structure of data. We explore different variable types, causality, and various data structures (Cross-sectional, Time Series, Panel). In this practical lab, we learn how to load and visualize these data structures using R and the `tidyverse` package.

# Introduction

In this first lab, we will load different data tables and visualize their structures as discussed in the theoretical presentation. In quantitative methods, it is essential to understand the type of data we are analyzing to apply the appropriate statistical analysis.

## Setting up the environment

We will start by loading the `tidyverse` package which provides a set of tools for data manipulation and visualization.

```{r setup, message=FALSE, warning=FALSE}
# Load necessary libraries
library(tidyverse)
```

## 1. Cross Sectional Data

Cross sectional data are observations of a sample of elements taken at one specific point in time. Let's look at a classic example: the `mtcars` dataset built into R.

```{r}
# Load cross-sectional data
data(mtcars)

# Visualize structure
glimpse(mtcars)

# Show first few rows
head(mtcars)
```

## 2. Time Series Data

Time series data consist of observations of one or more variables over time. Here we use the `economics` dataset from ggplot2.

```{r}
# Load time series data
data(economics)

# Visualize structure
glimpse(economics)

# Plotting time series structure quickly
ggplot(economics, aes(x = date, y = psavert)) +
  geom_line() +
  labs(title = "Personal Savings Rate Over Time", x = "Date", y = "Savings Rate")
```

## 3. Panel or Longitudinal Data

Panel data follows the same elements over multiple time periods. While base R doesn't have a perfect built-in panel dataset, we can simulate a simple structure to understand how it looks.

```{r}
# Creating a simple mock panel dataset
set.seed(123)
panel_data <- tibble(
  individual_id = rep(1:5, each = 3),
  year = rep(2021:2023, times = 5),
  income = round(runif(15, 30000, 80000), 0)
)

# Visualize structure
glimpse(panel_data)

# Show first few rows
head(panel_data)
```

# Session Information

Providing the session information ensures reproducibility by recording the exact versions of R and packages used in this analysis.

```{r session-info}
# Display session info
sessionInfo()
```

> [!tip] Cross-References
>
> * Return to the [Course Root](../../README.qmd)

> [!note] References and Materials
>
> **Download (Originals)**
> 
> - [W1. Intro and data.pptx](./W1_intro_and_data.pptx)
